---
title: "Reproductible Reseach: Peer Assessment 1"
author: "Ric"
date: "Saturday, August 16, 2014"
output: html_document
---

## Introduction

This is the first peer assessment for the Reproducible Research course from John Hopkins University, via coursera.  This assignment combines the tools of R programming, data analysis, knitr, git, gitHub, and other tools learned through the first five courses in this Data Science certificate 
program.    

All the R code and results will be displayed.   

Each major section in this assignment will be defined by a heading.  

## Loading and preprocessing the data

For this first section, we are to show how the raw code was imported into R for analysis and 
the steps proformed.  

The assumption is the raw csv file is in the working directory.  

```{r preprocess, echo=TRUE}
setwd("C:/Users/ric/RepData_PA1")
getwd() ## check working directory has been set. 
activityDataRaw <- read.csv("activity.csv")  ## read data into R dataframe. 
```
==============================  
Begin data file inspection  
==============================    
```{r}
str(activityDataRaw) ## inspect structure of data
summary(activityDataRaw)
head(activityDataRaw)  ## display first few rows of data
tail(activityDataRaw)  ## display last rows of data
sum(is.na(activityDataRaw$steps)) #  determine number of NA values (which is also seen in the summary)
```

```{r}
plot(activityDataRaw$date, activityDataRaw$steps)  ## quick look at two variables
```  

This completes the preprocessing section of this assignment.  


## What is the mean total number of steps taken per day?    

In this section of the assignment, the missing values (NA) observations are removed from the dataset.   

```{r}
activityData <- subset(na.omit(activityDataRaw))  ## build new dataset without NA values. 
head(activityData)  ## check first few rows.  
```   

Earlier, it was determined that there were 2304 NA observations in the original dataset of 17,568 observations.  Therefore, an arithmetic calculation shows that 13 percent of the original dataset has been removed.  

```{r}
2304/17568*100
```   
===================================================================   
Will now build a histogram of the total number of steps taken each day.  
===================================================================  
Calculate the total steps taken per day: 

```{r}
totalSteps <- tapply(activityData$steps, activityData$date, sum)  ## find the total steps count.
head(totalSteps)  ## show some of the data
```     

NA values have reappeared; therefore, the data will be sorted to move the NA values to the bottom and then removed:   

```{r}
totalStepsSorted <- sort(totalSteps, na.last=TRUE)  ## sort and move NA values to last
totalStepsData <- totalStepsSorted[c(1:53)]
```    

The resulting data:   

```{r}
totalStepsData  
```   

With this information, a basic historgram can be constructed:   

```{r}
hist(totalStepsData, breaks=20, xlab="Total number of steps taken", 
     main="Histogram of Total Steps Taken Each Day")   
```    

The **mean** and **median** of the total number of steps taken per day can be calculated with the R summary() function:    

```{r}
summary(totalStepsData)
```    

The **mean** is sometimes referred to as the average number, and the **median** is the number between the minimum and maximum values of the quantile.      

*****   
## What is the average daily activity pattern?    

For this section of the assignment, the **ggplot2** graphics system will be used.  This package is not part of the default R installation; therefore you may need to install the package.  If so, remove the comments characters from the following R code:   
```{r}
## install.packages("ggplot2")
```    
After the installation, the graphics package will need to be loaded:   
```{r}
library(ggplot2)
```   
The following R code will find the average number of steps for each five-minute interval: 
```{r}
avgSteps <- tapply(activityData$steps, activityData$interval, mean)
```   
This will result in the following data, showing the average number of steps across all days for the first five-minute interval, the second five-minute interval across all days for that interval labeled as 5, the third five-minute interval labeled as 10, and then 15, 20, etc.:  
```{r}
head(avgSteps)
```    
Convert the array into a data frame in preparation to plot the data:   
```{r}   
avgStepsDF <- as.data.frame(avgSteps)
```      

This data can be plotted with the following R code:   
```{r}
plot(avgStepsDF, type="l", xlab="Interval Position", ylab="Avg Daily Activity",
     main="Average Daily Activity Pattern")
```    

A visual inspection of the data shows that row 835 in the avgStepsDF contains the maximum number of steps, which is 206.1698.   
```{r}
max(avgStepsDF)
```     

## Inputting Missing Values    

Earlier, it was shown the number of missing values, coded as NA, from the original dataset.  The R code to determine the missing values is repeated here:   

```{r}
sum(is.na(activityDataRaw$steps)) #  determine number of NA values (which is also seen in the summary)
```      
As the output shows, there were 2304 observations that contained a NA value.    

The summary function would also show the NA values:    
```{r}
summary(activityDataRaw)    
```    
One way to fill in the missing values in the original dataset is with a For loop.  This For loop idea is from the [stackoverflow] (http://stackoverflow.com/questions/9322773/how-to-replace-na-with-mean-by-subset-in-r-impute-with-plyr) website.    
```{r}
activityDataModified <- activityDataRaw ## create a new dataset
## use a For loop to replace the NA values with the average value: 
for (i in which(sapply(activityDataRaw, is.numeric))) {
        activityDataModified[is.na(activityDataRaw[, i]), i] <- mean(activityDataRaw[, i],  na.rm = TRUE)
}
```      
A review of the new dataset now shows no NA values, being replaced with an average value: 
```{r}
summary(activityDataModified)
head(activityDataModified)  ## display first few rows of data
tail(activityDataModified)  ## display last rows of data
sum(is.na(activityDataModified$steps)) #  determine number of NA values (which is also seen in the summary)
```     
A histogram can be constructed on this new dataset:   
```{r}
totalStepsModified <- tapply(activityDataModified$steps, activityDataModified$date, sum)
## take the log() to better view the data:  
hist(log(activityDataModified$steps), xlab="Total number of steps taken",
        main="Revised Historgram of Total Steps Taken Each Day")
```       

The ***mean*** and ***median*** values of this new dataset are as follows:  
```{r}
summary(totalStepsModified)   
```    

Comparing this histogram with the histogram from the earlier analysis shows a possible pattern in the data.  The scales are different, but the earlier histogram does not have an easily recognizable pattern emerging as this newly built histogram shows.  Earlier, I mentioned that 13 percent of the dataset was not used because of the missing data.  However, having each NA value occurrence replaced with an average value resulted in a histogram suggesting a pattern where the peak quantity of steps occurred.  From an earlier analysis, this peak time was determined to be 835 which can be interrpreted from the data as 8:35 AM.      

## Are there differences in activity patterns between weekdays and weekends?    

This section of the assignment has not been completed due to time constraints.   

===========================    
This is the end of the document.   
===========================    
