---
title: "Reproductible Reseach: Peer Assessment 1"
author: "Ric"
date: "Saturday, August 16, 2014"
output: html_document
---

## Introduction

This is the first peer assessment for the Reproducible Research course from John Hopkins University, via coursera.  This assignment combines the tools of R programming, data analysis, knitr, git, gitHub, and other tools learned through the first five courses in this Data Science certificate 
program.    

All the R code and results will be displayed.   

Each section in this assignment will be defined by a heading.  

## Loading and preprocessing the data

For this first section, we are to show how the raw code was imported into R for analysis and 
the steps proformed.  

The assumption is the raw csv file is in the working directory.  

```{r preprocess, echo=TRUE}
setwd("C:/Users/ric/RepData_PA1")
getwd() ## check working directory has been set. 
activityDataRaw <- read.csv("activity.csv")  ## read data into R dataframe. 
```
==============================  
Begin data file inspection  
==============================    
```{r}
str(activityDataRaw) ## inspect structure of data
summary(activityDataRaw)
head(activityDataRaw)  ## display first few rows of data
tail(activityDataRaw)  ## display last rows of data
sum(is.na(activityDataRaw$steps)) #  determine number of NA values (which is also seen in the summary)
```

```{r}
plot(activityDataRaw$date, activityDataRaw$steps)  ## quick look at two variables
```  

This completes the preprocessing section of this assignment.  


## What is the mean total number of steps taken per day?    

In this section of the assignment, the missing values (NA) observations are removed from the dataset.   

```{r}
activityData <- subset(na.omit(activityDataRaw))  ## build new dataset without NA values. 
head(activityData)  ## check first few rows.  
```   

Earlier, it was determined that there were 2304 NA observations in the original dataset of 17,568 observations.  Therefore, an arithmetic calculation shows that 13 percent of the original dataset has been removed.  

```{r}
2304/17568*100
```   
===================================================================   
Will now build a historgram of the total number of steps taken each day.  
===================================================================  
Calculate the total steps taken per day: 

```{r}
totalSteps <- tapply(activityData$steps, activityData$date, sum)  ## find the total steps count.
head(totalSteps)  ## show some of the data
```     

NA values have reappeared; therefore, the data will be sorted to move the NA values to the bottom and then removed:   

```{r}
totalStepsSorted <- sort(totalSteps, na.last=TRUE)  ## sort and move NA values to last
totalStepsData <- totalStepsSorted[c(1:53)]
```    

The resulting data:   

```{r}
totalStepsData  
```   

With this information, a basic historgram can be constructed:   

```{r}
hist(totalStepsData, breaks=20, xlab="Total number of steps taken", 
     main="Histogram of Total Steps Taken Each Day")   
```    

The **mean** and **median** of the total number of steps taken per day can be calculated with the R summary() function:    

```{r}
summary(totalStepsData)
```    

The mean is sometimes referred to as the average number, and the median is the number between the minimum and maximum values of the quantile.      

*****   
## What is the average daily activity pattern?    

